### 2020-09-21 paper/coloringBergeGraphs.tex
Overweekend fixes

There has been much progress on the quest of finding an algorithm coloring perfect graphs, without the use of ellipsoid method (see \Cref{sec:classicalColoring}), however there is still no known polynomial combinatorial algorithm to do this.

@@

From (\ref{en:ei2}) it follows that diagonal entries of any positive semidefinite matrix are nonnegative and the sum of two positive semidefinite matrices is positive semidefinite.

@@

Now, we can define a \emph{semidefinite program} to be an optimization problem of the following form:

\begin{equation*}
  \begin{array}{ll@{}ll}
    \text{minimize}   & c^\intercal x                          & \\
    \text{subject to} & x_1A_1 + \ldots + x_nA_n - B \succeq 0   \\
  \end{array}
\end{equation*}
Here $A_1, \ldots, A_n, B$ are given symmetric $m \times m$ matrices and $c \in \mathbb{R}^n$ is a given vector. Any choice of the values $x_i$ that satisfies the given constraint is called a \emph{feasible solution}.

@@

Let us recall, that an orthogonal representation of a graph $G = (V, E)$ is a labeling $u: V \rightarrow \mathbb{R}^d$ for some $d$, such that $u_i^\intercal u_j = 0$ for all anticonnected vertices $i, j$. An \emph{orthonormal} representation is an orthogonal representation with $|u_i| = 1$ for all $i$. The \emph{angle} of an orthogonal representation is the smallest half-angle of a rotational cone containing the representing vectors.

@@

  \begin{equation*}
    \begin{array}{ll@{}ll}
      \text{minimize}   & t      &                                              \\
      \text{subject to} & Y      & \succeq 0                                    \\
                        & Y_{ij} & = -1      & (\forall~ij \in E(\overline{G})) \\
                        & Y_{ii} & = t - 1                                      \\
    \end{array}
  \end{equation*}

@@

\section{Coloring perfect graphs using ellipsoid method}
\label{sec:coloringEllipsoid}
The following part is based on \citetitle{Laurent2005} by \citeauthor{Laurent2005} \cite{Laurent2005}.

@@

Ever since Grötschel et al. published an ellipsoid-method-based polynomial algorithm for coloring perfect graphs, a combinatorial algorithm for doing the same has been sought. As of yet, it is not known, although there is much progress in the field.

\begin{defn}[prism]
  A \emph{prism} is a graph consisting of two disjoint triangles and two disjoint paths between them. Notice that for a graph to contain no odd holes, all three paths in a prism must have the same parity.
  \label{def:prism}
\end{defn}

A prism with all three paths odd is called an \emph{odd prism} and a prism with all three paths even is called an \emph{even prism}.

@@

A most recent advancement (2018) is an algorithm by Chudnovsky et al. that colors any square-free Berge graphs in time of $O(|V|^9)$ \cite{coloringSquareFree}. Before proving strong perfect graph conjecture a similar conjecture for square-free Berge graphs has been proven by Conforti et al. \cite{Conforti2004}. During one of her lectures, Maria Chudnovsky expressed hope that discovery of full algorithm for coloring Berge graphs might follow a similar pattern.

The algorithm for coloring square-free Berge graphs is much more complex than CCLSV algorithm for Berge graphs recognition and would be much harder to implement. The main idea is that, if square-free Berge graph $G$ is not a member of class that is simple to recognize and color (Artemis graphs), then it must have an induced prism. We find it and grow it until it is maximal and then we can create a partition of $G$, that we color recursively. The paper in which the algorithm is presented \cite{coloringSquareFree} is of much more mathematical ''flavor'' than \cite{MC05} -- there is a proof that a $O(|V|^9)$ algorithm exists, but actually deducing it from the paper is far from straightforward. We provide its pseudocode in the Appendix A.

>>>>

There has been much progress on the quest of finding an algorithm coloring perfect graphs, without the use of the ellipsoid method (see \Cref{sec:classicalColoring}), however there is still no known polynomial combinatorial algorithm to do this.

@@

From (\ref{en:ei2}) it follows that diagonal entries of any positive semidefinite matrix are nonnegative and that the sum of two positive semidefinite matrices is positive semidefinite.

@@

\begin{defn}[semidefinite program]
  Let $A_1, \ldots, A_n, B$ be given symmetric $m \times m$ matrices and be $c \in \mathbb{R}^n$ a given vector. A \emph{semidefinite program} is an optimization problem of the following form:
  \begin{equation*}
    \begin{array}{ll@{}ll}
      \text{minimize}   & c^\intercal x                          & \\
      \text{subject to} & x_1A_1 + \ldots + x_nA_n - B \succeq 0   \\
    \end{array}
  \end{equation*}
  Any choice of the values $x_i$ that satisfies the given constraint is called a \emph{feasible solution}.
\end{defn}

@@

Let us recall, that an orthogonal representation of a graph $G = (V, E)$ is a labeling $u: V \rightarrow \mathbb{R}^d$ for some $d$, such that $u_i^\intercal u_j = 0$ for all anticonnected vertices $i, j$. The \emph{angle} of an orthogonal representation is the smallest half-angle of a rotational cone containing the representing vectors.

@@

  \begin{equation*}
    \begin{array}{l@{}rlll}
      \text{minimize~~}   & t      &                                              \\
      \text{subject to~~} & Y      & \succeq & 0                                    \\
                        & Y_{ij} & = & -1      & (\forall~ij \in E(\overline{G})) \\
                        & Y_{ii} & = & t - 1                                      \\
    \end{array}
  \end{equation*}

@@

\section{Coloring perfect graphs using an ellipsoid method}
\label{sec:coloringEllipsoid}
The following part is based on \citetitle{Laurent2005} by \citeauthor{Laurent2005} \cite{Laurent2005}.

@@

Ever since Grötschel et al. published an ellipsoid-method-based polynomial algorithm for coloring perfect graphs, a combinatorial algorithm for doing the same has been sought. As of yet, there isn't one, although there is much progress in the field.

A \emph{prism} is a graph consisting of two disjoint triangles and two disjoint paths between them. Notice that for a graph to contain no odd holes, all three paths in a prism must have the same parity. A prism with all three paths odd is called an \emph{odd prism} and a prism with all three paths even is called an \emph{even prism}.

@@

A most recent advancement (2018) is an algorithm by Chudnovsky et al. that colors any square-free Berge graphs in time of $O(|V|^9)$ \cite{coloringSquareFree}. Before proving strong perfect graph conjecture a similar conjecture for square-free Berge graphs has been proven by Conforti et al. \cite{Conforti2004}. During one of her lectures, Maria Chudnovsky expressed hope that discovery of a full algorithm for coloring Berge graphs might follow a similar pattern.

The algorithm for coloring square-free Berge graphs is much more complex than CCLSV algorithm for Berge graphs recognition and would be much harder to implement. The main idea is that, if square-free Berge graph $G$ is not a member of class that is simple to recognize and color (Artemis graphs), then it must have an induced prism. We find it and grow it until it is maximal and then we can create a partition of $G$, that we color recursively. The paper in which the algorithm is presented \cite{coloringSquareFree} is of much more mathematical ''flavor'' than \cite{MC05} -- there is a proof that a $O(|V|^9)$ algorithm exists, but actually deducing it from the paper isn't straightforward. We provide its pseudocode in the Appendix A.

### 2020-09-21 paper/definitions.tex
Overweekend fixes

We will only consider finite graphs, that is, $V$ and $E$ are always finite. If $G$ is a graph, then $V = V(G)$ is the \emph{vertex set} of $G$, and $E = E(G)$ is the \emph{edge set}. When the context of $G$ is clear we will use $V$ and $E$ to denote its vertex and edge set.

@@

Given a graph $G = (V, E)$ and a set $X \subseteq V$ by $G\setminus X$ we will denote a induced subgraph $G[V\setminus X]$.

@@

Our graph $G_0$ is a path of length 3, with the inside $G_0^* = \{v_2, v_3\}$. If we  added any edge to $G_0$ it would stop being a path.

@@

\begin{defn}[clique]
  A \emph{complete graph} or a \emph{clique} is a graph of the form $G = (V, {V \choose 2})$, that is, every two vertices are connected. We will denote a clique on $n$ vertices as $K_n$.
\end{defn}

\begin{defn}[clique number]
  A \emph{clique number} of a graph $G$, denoted as $\omega(G)$, is a cardinality of its largest induced clique.
\end{defn}

@@

\begin{defn}[stability number]
  A \emph{stability number} of a graph $G$, denoted as $\alpha(G)$, is a cardinality of its largest induced stable set.
\end{defn}

@@

\begin{defn}[chromatic number]
  A \emph{chromatic number} of a graph $G$, denoted as $\chi(G)$, is a smallest natural number $k$, for which there exists a $k$-coloring of $G$.
\end{defn}

>>>>

We will only consider finite graphs, that is, $V$ and $E$ are always finite. If $G$ is a graph, then $V = V(G)$ is the \emph{vertex set} of $G$, and $E = E(G)$ is the \emph{edge set} of $G$. When the context of $G$ is clear we will use $V$ and $E$ to denote its vertex and edge set.

@@

Given a graph $G = (V, E)$ and a set $X \subseteq V$ by $G\setminus X$ we will denote the induced subgraph $G[V\setminus X]$.

@@

Our graph $G_0$ is a path of length 3, with the inside $G_0^* = \{v_2, v_3\}$. If we added any edge to the $G_0$ it would stop being a path.

@@

\begin{defn}[clique]
  A \emph{complete graph} or a \emph{clique} is a graph of the form $G = (V, {V \choose 2})$, that is, every two vertices are connected.
\end{defn}

We will denote the clique on $n$ vertices as $K_n$.

\begin{defn}[clique number]
  The \emph{clique number} of a graph $G$, denoted as $\omega(G)$, is a cardinality of its largest induced clique.
\end{defn}

@@

\begin{defn}[stability number]
  The \emph{stability number} of a graph $G$, denoted as $\alpha(G)$, is a cardinality of its largest induced stable set.
\end{defn}

@@

\begin{defn}[chromatic number]
  The \emph{chromatic number} of a graph $G$, denoted as $\chi(G)$, is a smallest natural number $k$, for which there exists a $k$-coloring of $G$.
\end{defn}

### 2020-09-21 paper/experiments.tex
Overweekend fixes

Examples of full binary graphs, and lattice graphs show that getting possible near cleaners still can be a big bottleneck. Parallelization of this method could be considered in the future. A possible idea would be to use recently developed dynamic dictionary for the GPU \cite{Ashkiani2018}. When running GPU CCLSV on split graphs, quite a lot of time is spent checking simple structures. The biggest challenge we can see is in an attempt to parallelize this is spreading work evenly across CUDA threads.

@@

Calculating $\vartheta(G)$ is the most complicated part of the coloring algorithm. With that done by an external library, the rest of the program is a straightforward implementation of the algorithms in \Cref{sec:coloringEllipsoid}. In most of our tests, the majority of running time was consumed on calculating $\vartheta$ of various graphs. As it was done by an external library, there isn't much optimization potential for us. One thing of note is, that by specifying the precision of $\vartheta$ we want to calculate to be $1/3$ the algorithm sped up  by 20-40\%. Also, we cache results of calculating $\vartheta$, meaning that through whole execution of our coloring algorithm, we only calculate $\vartheta$ for each graph once. Our main goal of the implementation was to check if this method is still impractical, even on modern equipment. Let us proceed straight to experiments and results.

@@

A GPU implementation of the na\"ive algorithm could also be considered for future work. We have created one, but after all path-enumeration optimizations of the CPU implementation we didn't manage to achieve any noticeable speedup. The reason again was, that we couldn't spread the work evenly between CUDA threads. Because some prefixes of paths require much more processing than others, this is not a trivial task and a careful study would be needed.

>>>>

Examples of full binary graphs, and lattice graphs show that getting possible near cleaners still can be a big bottleneck. Parallelization of this method could be considered in the future. A possible idea would be to use recently developed dynamic dictionary for the GPU \cite{Ashkiani2018}. When running GPU CCLSV on split graphs, quite a lot of time is spent checking simple structures. The biggest challenge we can see is in an attempt to parallelize this is spreading the work evenly across CUDA threads.

@@

Calculating $\vartheta(G)$ is the most complicated part of the coloring algorithm. With that done by an external library, the rest of the program is a straightforward implementation of the algorithms in \Cref{sec:coloringEllipsoid}. In most of our tests, the majority of running time was consumed on calculating $\vartheta$ of various graphs. So, there isn't much optimization potential for us. One thing of note is, that by specifying the precision of $\vartheta$ we want to calculate to be $1/3$ the algorithm sped up  by 20-40\%. Also, we cache results of calculating $\vartheta$, meaning that through whole execution of our coloring algorithm, we only calculate $\vartheta$ for each graph once. Our main goal of the implementation was to check if this method is still impractical, even on modern equipment. Let us proceed straight to experiments and results.

@@

A GPU implementation of the na\"ive algorithm could also be considered for a future work. We have created one, but after all path-enumeration optimizations of the CPU implementation we didn't manage to achieve any noticeable speedup. The reason again was, that we couldn't spread the work evenly between CUDA threads. Because some prefixes of paths require much more processing than others, this is not a trivial task and a careful study would be needed.

### 2020-09-21 paper/implementation.tex
Overweekend fixes

We try find an odd hole directly, then, we run the same algorithm on $\overline{G}$, and if no odd hole is found we report that the graph is perfect, else that it is not.

@@

Another crucial optimization is to eliminate partial solutions that have edges between not consecutive vertices right away. Also, when we look for an odd hole specifically, we can assume that its first vertex we enumerate is the lowest. Those two optimizations make a huge difference. With them, our na\"ive algorithm implementation went from being able to check graphs with $|V|$ up to 12-14, within a reasonable time, to being faster than CCLSV on most of our tests.

>>>>

We try find an odd hole directly, then, we run the same algorithm on $\overline{G}$, and if no odd holes are found we report that the graph is perfect, else that it is not.

@@

Another crucial optimization is to eliminate partial solutions that have edges between nonconsecutive vertices right away. Also, when we look for an odd hole specifically, we can assume that its first vertex we enumerate is the lowest. Those two optimizations make a huge difference. With them, our na\"ive algorithm implementation went from being able to check graphs with $|V|$ up to 12-14, within a reasonable time, to being faster than CCLSV on most of our tests.

### 2020-09-21 paper/perfectGraphs.tex
Overweekend fixes

Given a graph $G$, let us consider a problem of coloring it using as few colors as possible. If $G$ contains a clique $K$ as a subgraph, we must use at least $|V(K)|$ colors to color it. This gives us a lower bound for a chromatic number $\chi(G)$ -- it is always greater or equal to the cardinality of the largest clique $\omega(G)$. The reverse is not always true, in fact we can construct a graph with no triangles and requiring arbitrarily large numbers of colors (e.g. construction by Mycielski \cite{Mycielski1955}).

@@

\begin{defn}[Berge graph]
	\label{def:bergeGraph}
	A graph $G$ is \emph{Berge} if and only if both $G$ and $\overline{G}$ have no odd hole as an induced subgraph.
\end{defn}

@@

Odd holes are not perfect, since their chromatic number is 3 and their largest cliques are of size 2. It is also easy to see, that an odd antihole of size $n$ has a chromatic number of $\frac{n+1}{2}$ and largest cliques of size $\frac{n-1}{2}$. A graph with no odd hole and no odd antihole is called \emph{Berge} (\Cref{def:bergeGraph}) after Claude Berge who studied perfect graphs.

In 1961 Berge conjured that a graph is perfect if and only if it contains no odd hole and no odd antihole in what has become known as a strong perfect graph conjecture. In 2001 Chudnovsky et al. have proven it and published the proof in an over 150 pages long paper \citetitle{MC06} \cite{MC06}.

@@

\section{Recognizing berge graphs}
\label{sec:recognizingBerge}

@@

Most obvious way to find a jewel is to enumerate all (possibly chordal) cycles of length 5 as $v_1, \ldots v_5$, check if it has all required nonedges and if it does, try to find a path $P$ as required. This gives us a time of $O(|V|^7)$. We could speed it up to $O(|V|^6)$ with more careful algorithm (see Algorithm 3.1 of \cite{MC05}), but since whole Berge recognition algorithm takes time $O(|V|^9)$ and our testing showed that time it takes to test for jewels is negligible we decided against it.

@@

A configuration of type $\T_3$ is a sequence $v_1, \ldots, v_6$, $P$, $X$, such that:
\begin{itemize}
	\item $v_1, \ldots v_6$ are distinct vertices,
	\item $v_1v_2$, $v_3v_4$, $v_1v_4$, $v_2v_3$, $v_3v_5$, $v_4v_6$ are edges, and $v_1v_3$, $v_2v_4$, $v_1v_5$, $v_2v_5$, $v_1v_6$, $v_2v_6$, $v_4v_5$ are nonedges,
	\item $X$ is an anticomponent of the set of all $\{v_1, v_2, v_5\}$-complete vertices, and $v_3$, $v_4$ are not $X$-complete,
	\item $P$ is a path of $G \setminus ( X \cup \{v_1, v_2, v_3, v_4\} )$ between $v_5$ and $v_6$ and no vertex in $P*$ is $X$-complete or adjacent to $v_1$ or adjacent to $v_2$,
	\item If $v_5v_6$ is an edge, then $v_6$ is not $X$-complete.
\end{itemize}

@@

	Each execution of lines \ref{line:t3n4start}-\ref{line:t3mid} takes $O(|V|^2)$ time, and there are $O(|V|)$ $v_4$s to enumerate. Line \ref{line:t3v4if} takes $O(|V|)$ time and then there are $O(|V|)$ $v_3$ to enumerate. Each is checked in $O(1)$ time. If we find the correct one, we calculate a path $P$, but return right away, so we calculate $P$ at most once in whole algorithm.
\end{proof}

@@

\begin{algorithmic}[1]
	\mProcedure{Test-Near-Cleaner}{$G, X$}
		\mForEach{pair $x, y \in V(G)$, such that $x \neq y$}
			\mIf{there is a path between $x$ and $y$ with no internal vertices in $X$ \lsx}
				\ls $R(x, y) \gets$ the shortest path $R(x, y)$ between $x, y$ with no
				\lsx internal vertex in $X$.
				\ls $r(x, y) \gets$ the length of $R(x, y)$
			\mElse
				\ls $R(x, y) \gets \textsc{undefined}$
				\ls $r(x, y) \gets \infty$
			\mEndIf
			\mForEach{$y_1 \in V(G) \setminus X$} \label{line:cudaStart}
				\mForEach{3-vertex path $x_1$-$x_2$-$x_3$ of $G\setminus y_1$} \label{line:3vertex}
					\mIf{$R(x_1, y_1), R(x_2, y_2)$ both exist} \label{line:cudaWorkStard}
						\ls $y_2 \gets$ the neighbor of $y_1$ in $R(x_2, y_1)$ \label{line:definey2}
						\mIf{$r(x_2, y_1) = r(x_1, y_1) + 1 = r(x_1, y_2)$ \AND
						\lsx $r(x_2, y_1) \leq min(r(x_3, y_1), r(x_3, y_2))$ \label{line:manyconditions}}
						\ls \RETURN an odd hole found
						\mEndIf
					\mEndIf \label{line:cudaWorkEnd}
				\mEndFor
			\mEndFor \label{line:cudaEnd}
		\mEndFor
		\ls \RETURN no odd hole found
	\mEndProcedure
\end{algorithmic}

@@

	The overall running time is $O(|V|^9)$ with testing for pyramids and testing all possible near-cleaners being parts with biggest time complexity.
\end{proof}

>>>>

Given a graph $G$, let us consider a problem of coloring it using as few colors as possible. If $G$ contains a clique $K$ as a subgraph, we must use at least $|V(K)|$ colors to color it. This gives us a lower bound for the chromatic number $\chi(G)$ -- it is always greater or equal to the cardinality of the largest clique $\omega(G)$. The reverse is not always true, in fact we can construct a graph with no triangles and requiring arbitrarily large numbers of colors (e.g. construction by Mycielski \cite{Mycielski1955}).

@@

\begin{defn}[Berge graph]
	\label{def:bergeGraph}
	A graph $G$ is \emph{Berge} if and only if both $G$ and $\overline{G}$ have no odd holes as induced subgraphs.
\end{defn}

@@

Odd holes are not perfect, since their chromatic number is 3 and their largest cliques are of size 2. It is also easy to see, that an odd antihole of size $n$ has the chromatic number of $\frac{n+1}{2}$ and largest cliques of size $\frac{n-1}{2}$. A graph with no odd holes and no odd antiholes is called \emph{Berge} (\Cref{def:bergeGraph}) after Claude Berge who studied perfect graphs.

In 1961 Berge conjured that a graph is perfect if and only if it contains no odd holes and no odd antiholes in what has become known as a strong perfect graph conjecture. In 2001 Chudnovsky et al. have proven it and published the proof in an over 150 pages long paper \citetitle{MC06} \cite{MC06}.

@@

\section{Recognizing perfect graphs}
\label{sec:recognizingBerge}

@@

The most obvious way to find a jewel is to enumerate all (possibly chordal) cycles of length 5 as $v_1, \ldots v_5$, check if it has all required nonedges and if it does, try to find a path $P$ as required. This gives us a time of $O(|V|^7)$. We could speed it up to $O(|V|^6)$ with a more careful algorithm (see Algorithm 3.1 of \cite{MC05}), but since whole Berge recognition algorithm takes time $O(|V|^9)$ and our testing showed that time it takes to test for jewels is negligible we decided against it.

@@

A configuration of type $\T_3$ is a sequence $v_1, \ldots, v_6$, $P$, $X$, such that:
\begin{itemize}
	\item $v_1, \ldots v_6$ are distinct vertices,
	\item $v_1v_2$, $v_3v_4$, $v_1v_4$, $v_2v_3$, $v_3v_5$, $v_4v_6$ are edges, and $v_1v_3$, $v_2v_4$, $v_1v_5$, $v_2v_5$, $v_1v_6$, $v_2v_6$, $v_4v_5$ are nonedges,
	\item $X$ is an anticomponent of the set of all $\{v_1, v_2, v_5\}$-complete vertices, and $v_3$, $v_4$ are not $X$-complete,
	\item $P$ is a path of $G \setminus ( X \cup \{v_1, v_2, v_3, v_4\} )$ between $v_5$ and $v_6$ and no vertex in $P^*$ is $X$-complete or adjacent to $v_1$ or adjacent to $v_2$,
	\item If $v_5v_6$ is an edge, then $v_6$ is not $X$-complete.
\end{itemize}

@@

	Each execution of lines \ref{line:t3n4start}-\ref{line:t3mid} takes $O(|V|^2)$ time, and there are $O(|V|)$ $v_4$s to enumerate. Line \ref{line:t3v4if} takes $O(|V|)$ time and then there are $O(|V|)$ $v_3$ to enumerate. Each is checked in $O(1)$ time. If we find the correct one, we calculate a path $P$, but return right away, so we calculate $P$ at most once in the whole algorithm.
\end{proof}

@@

\begin{algorithmic}[1]
	\mProcedure{Test-Near-Cleaner}{$G, X$}
		\mForEach{pair $x, y \in V(G)$, such that $x \neq y$}
			\mIf{there is a path between $x$ and $y$ with no internal vertices in $X$ \lsx}
				\ls $R(x, y) \gets$ the shortest path $R(x, y)$ between $x, y$ with no
				\lsx internal vertex in $X$.
				\ls $r(x, y) \gets$ the length of $R(x, y)$
			\mElse
				\ls $R(x, y) \gets \textsc{undefined}$
				\ls $r(x, y) \gets \infty$
			\mEndIf
			\mForEach{$y_1 \in V(G) \setminus X$} \label{line:cudaStart}
				\mForEach{3-vertex path $x_1$-$x_2$-$x_3$ of $G\setminus y_1$} \label{line:3vertex}
					\mIf{$R(x_1, y_1), R(x_2, y_2)$ both exist} \label{line:cudaWorkStard}
						\ls $y_2 \gets$ the neighbor of $y_1$ in $R(x_2, y_1)$ \label{line:definey2}
						\mIf{$r(x_2, y_1) = r(x_1, y_1) + 1 = r(x_1, y_2)$ \AND
						\lsx $r(x_2, y_1) \leq min(r(x_3, y_1), r(x_3, y_2))$ \label{line:manyconditions}}
						\ls \RETURN an odd hole found
						\mEndIf
					\mEndIf \label{line:cudaWorkEnd}
				\mEndFor
			\mEndFor \label{line:cudaEnd}
		\mEndFor
		\ls \RETURN no odd holes found
	\mEndProcedure
\end{algorithmic}

@@

	The overall running time is $O(|V|^9)$ with testing for pyramids and testing all possible near-cleaners being parts with the biggest time complexity.
\end{proof}
